Reducing bias in secondary data analysis via an Explore and Confirm Analysis Workflow (ECAW): A proposal and survey of observational researchers

### 2023-04-26 manuscript/manuscript.Rmd
finished converting the manuscript and cleaning up the codes

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
library("papaja")
library(tidyverse)
library(here)
library(kableExtra)
library(gt)
library(patchwork)
# The developer version is needed
library(likert)
library(knitr)
library(viridis)
r_refs("ecaw-ref.bib")
```

@@

```{r loadingData}
processed <- read_csv(here::here("data/processed/ecaw_processed_data.csv"))
processed_all <- read_csv(here::here("data/processed/ecaw_processed_all_data.csv"))
```

```{r analysisPreferences}
# Seed for random number generation
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

@@

# Results
## Survey completion
Of 1148 emails sent, 1094 went through and 54 bounced. The survey was completed `r nrow(processed)` times and partially completed `r nrow(dplyr::filter(processed, has_missing)) ` times, leading to a response rate of `r round(nrow(processed) / 1094 * 100)`% for complete surveys and `r round(nrow(filter(processed, has_missing)) / 1094 * 100)`% for incomplete surveys.\footnote[1]{The ALSPAC mailing list has been maintained as a record of collaborators for many years and is constantly added to. The ALSPAC team only removes email addresses from their mailing list if someone explicitly requests this action. Thus, their list likely includes several email addresses that are no longer monitored. For example, we received one email reply stating that the recipient hasn’t been active in research for 30 years. It is also possible that some researchers have multiple email addresses on the mailing list (e.g., because they moved institutions). These two factors may have deflated the response rate.} The median time taken for complete survey responses was `r round(median(processed$duration_in_mins), 1)` minutes (IQR: `r round(quantile(processed$duration_in_mins, .25, na.rm = TRUE), 1)` to `r round(quantile(processed$duration_in_mins, .75, na.rm = T), 1)`). This manuscript presents the results for complete surveys. Supplementary Material D presents the results with partially complete surveys included.

@@

Respondents had published a median of `r median(processed$n_studies)` (IQR `r round(quantile(processed$n_studies, .25, na.rm = TRUE), 1)` to `r round(quantile(processed$n_studies, .75, na.rm = TRUE), 1)`) studies using preexisting observational data (Supplementary Figure C1). They reported using the following programming languages or software packages: `r language_text` (Supplementary Table C1)\footnote[2]{Participants could select multiple responses to this survey question.}. `r pull(filter(concerned, support == "positive"), percentage)`% (`r pull(filter(concerned, support == "positive"), n)`/`r pull(filter(concerned, support == "positive"), n_sum)`) of participants reported being more concerned with research trustworthiness, bias, rigour, and reproducibility compared to what they think of as a typical research who uses preexisting observational data; `r pull(filter(concerned, support == "negative"), percentage)`% (`r pull(filter(concerned, support == "negative"), n)`/`r pull(filter(concerned, support == "negative"), n_sum)`) reported being less concerned (Supplementary Figure C2).

@@

(ref:typicallyEcawPlotCaption) __Figure 1. Responses to the survey questions on trustworthiness and reproducibility of observational research with preexisting data and ECAWs.__ The survey defined trustworthy as “meaning that the results and conclusions of the publications are valid, reliable, rigorous, and accurate. That they merit trust”. The survey defined reproducible “in the sense that other researchers re-analysing the data with the same research question would produce similar results.” For each item, the number to the left of the data bar indicates the combined percentage for the responses depicted in any shade of brown/orange. The number in the center of the data bar (gray) indicates the percentage of neutral responses. The number to the right of the data bar indicates the combined percentage for the responses depicted in any shade of green. The bar charts in the top panel had no missing responses or selection of the option _“I don’t understand the question”_. The bar charts in the bottom panel excluded missing responses (n = `r pull(filter(ecaw_missing, item == "Trustworthy" & value == "Missing"), n)`; `r pull(filter(ecaw_missing, item == "Reproducible" & value == "Missing"), n)`) and responses of _“I don't understand the question”_ (n = `r pull(filter(ecaw_missing, item == "Trustworthy" & value == "I don't understand the question"), n)`; `r pull(filter(ecaw_missing, item == "Reproducible" & value == "I don't understand the question"), n)`).

@@

# method_missing_data <-
#   method_data %>%
#   filter(scale %in% c("Missing", "I don't understand the question")) %>%
#   mutate(
#     scale = case_when(
#       scale == "I don't understand the question" ~ "Don't understand",
#       scale == "Missing" ~ "Missing"),
#     text = paste0(scale, ": ", n)
#     ) %>%
#   group_by(type) %>%
#   summarise(text = stringr::str_c(text, collapse = "\n"))

@@

(ref:alspacPlotCaption) __Figure 3. Responses to survey questions about using ECAWs.__ These bar charts exclude missing values (_n_ = `r pull(filter(alspac_missing, item == "Less_willing" & value == "Missing"), n)`; `r pull(filter(alspac_missing, item == "Opt_in" & value == "Missing"), n)`; `r pull(filter(alspac_missing, item == "Study" & value == "Missing"), n)`; `r pull(filter(alspac_missing, item == "Prefer_ecaw" & value == "Missing"), n)`; respectively from top to bottom), responses of _"I don't understand the question"_ (_n_ = `r pull(filter(alspac_missing, item == "Less_willing" & value == "I don't understand the question"), n)`; `r pull(filter(alspac_missing, item == "Opt_in" & value == "I don't understand the question"), n)`; `r pull(filter(alspac_missing, item == "Study" & value == "I don't understand the question"), n)`; `r pull(filter(alspac_missing, item == "Prefer_ecaw" & value == "I don't understand the question"), n)`), and responses of _"Unsure"_ (_n_ = `r pull(filter(alspac_missing, item == "Less_willing" & value == "Unsure"), n)`; `r pull(filter(alspac_missing, item == "Opt_in" & value == "Unsure"), n)`; `r pull(filter(alspac_missing, item == "Study" & value == "Unsure"), n)`; `r pull(filter(alspac_missing, item == "Prefer_ecaw" & value == "Unsure"), n)`). Agreement with the first question may be slightly inflated due to the format of the questions in this block. Respondents with a highly positive inclination towards ECAWs would be expected to disagree with the first question, but agree with the next three questions. Four respondents agreed with all four statements, suggesting they may have glazed over the word “less” in the first question \protect\footnotemark[3]. Interpreting responses to the second and third question come with a degree of ambiguity as the survey did not specify what was meant by the term “study” \protect\footnotemark[4].

>>>>

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
# Used for rendering the manuscript rmarkdown document
library("papaja")
# Used for general data wrangling
library(tidyverse)
# Used to make exact file paths
library(here)
# Used for tables
library(kableExtra)
library(gt)
# Used for joining figures
library(patchwork)
# The developer version is needed
# Used for creating Likert ggplots
library(likert)
# Used for knitting rmarkdown documents
library(knitr)
# Used for colorblind ggplot colors
library(viridis)
# References
r_refs("ecaw-ref.bib")
```

@@

```{r loadingData}
# Loading processed datafiles
processed <- read_csv(here::here("data/processed/ecaw_processed_data.csv"))
processed_all <- read_csv(here::here("data/processed/ecaw_processed_all_data.csv"))
```

@@

# Results
## Survey completion
Of 1148 emails sent, 1094 went through and 54 bounced. The survey was completed `r nrow(processed)` times and partially completed `r nrow(processed_all) - nrow(processed)` times, leading to a response rate of `r round(nrow(processed) / 1094 * 100)`% for complete surveys and `r round(nrow(processed_all) / 1094 * 100)`% for at least partially complete surveys.\footnote[1]{The ALSPAC mailing list has been maintained as a record of collaborators for many years and is constantly added to. The ALSPAC team only removes email addresses from their mailing list if someone explicitly requests this action. Thus, their list likely includes several email addresses that are no longer monitored. For example, we received one email reply stating that the recipient hasn’t been active in research for 30 years. It is also possible that some researchers have multiple email addresses on the mailing list (e.g., because they moved institutions). These two factors may have deflated the response rate.} The median time to complete the survey was `r round(median(processed$duration_in_mins), 1)` minutes (IQR: `r round(quantile(processed$duration_in_mins, .25, na.rm = TRUE), 1)` to `r round(quantile(processed$duration_in_mins, .75, na.rm = T), 1)`). This manuscript presents the results for complete surveys. Supplementary Material D presents the results with partially complete surveys included.

@@

Respondents had published a median of `r median(processed$n_studies, na.rm = TRUE)` (IQR `r round(quantile(processed$n_studies, .25, na.rm = TRUE))` to `r round(quantile(processed$n_studies, .75, na.rm = TRUE))`) studies using preexisting observational data (Supplementary Figure C1). They reported using the following programming languages or software packages: `r language_text` (Supplementary Table C1)\footnote[2]{Participants could select multiple responses to this survey question.}. `r pull(filter(concerned, support == "positive"), percentage)`% (`r pull(filter(concerned, support == "positive"), n)`/`r pull(filter(concerned, support == "positive"), n_sum)`) of participants reported being more concerned with research trustworthiness, bias, rigour, and reproducibility compared to what they think of as a typical research who uses preexisting observational data; `r pull(filter(concerned, support == "negative"), percentage)`% (`r pull(filter(concerned, support == "negative"), n)`/`r pull(filter(concerned, support == "negative"), n_sum)`) reported being less concerned (Supplementary Figure C2).

@@

(ref:typicallyEcawPlotCaption) __Figure 1. Responses to the survey questions on trustworthiness and reproducibility of observational research with preexisting data and ECAWs.__ The survey defined trustworthy as “meaning that the results and conclusions of the publications are valid, reliable, rigorous, and accurate. That they merit trust”. The survey defined reproducible “in the sense that other researchers re-analysing the data with the same research question would produce similar results.” For each item, the number to the left of the data bar indicates the combined percentage for the responses depicted in any shade of brown/orange. The number in the center of the data bar (gray) indicates the percentage of neutral responses. The number to the right of the data bar indicates the combined percentage for the responses depicted in any shade of green. The bar charts in the top panel had no missing responses or selection of the option _“I don’t understand the question”_. The bar charts in the bottom panel excluded responses of _“I don't understand the question”_ (n = `r pull(filter(ecaw_missing, item == "Trustworthy"), n)`; `r pull(filter(ecaw_missing, item == "Reproducible"), n)`).

@@

@@

(ref:alspacPlotCaption) __Figure 3. Responses to survey questions about using ECAWs.__ These bar charts exclude responses of _"I don't understand the question"_ (_n_ = `r pull(filter(alspac_missing, item == "Less_willing" & value == "I don't understand the question"), n)`; `r pull(filter(alspac_missing, item == "Opt_in" & value == "I don't understand the question"), n)`; `r pull(filter(alspac_missing, item == "Study" & value == "I don't understand the question"), n)`; `r pull(filter(alspac_missing, item == "Prefer_ecaw" & value == "I don't understand the question"), n)`; respectively from top to bottom), and responses of _"Unsure"_ (_n_ = `r pull(filter(alspac_missing, item == "Less_willing" & value == "Unsure"), n)`; `r pull(filter(alspac_missing, item == "Opt_in" & value == "Unsure"), n)`; `r pull(filter(alspac_missing, item == "Study" & value == "Unsure"), n)`; `r pull(filter(alspac_missing, item == "Prefer_ecaw" & value == "Unsure"), n)`). Agreement with the first question may be slightly inflated due to the format of the questions in this block. Respondents with a highly positive inclination towards ECAWs would be expected to disagree with the first question, but agree with the next three questions. Four respondents agreed with all four statements, suggesting they may have glazed over the word “less” in the first question \protect\footnotemark[3]. Interpreting responses to the second and third question come with a degree of ambiguity as the survey did not specify what was meant by the term “study” \protect\footnotemark[4].

### 2023-04-26 manuscript/manuscript.tex
finished converting the manuscript and cleaning up the codes

Of 1148 emails sent, 1094 went through and 54 bounced. The survey was completed 103 times and partially completed 17 times, leading to a response rate of 9\% for complete surveys and 2\% for incomplete surveys.\footnote[1]{The ALSPAC mailing list has been maintained as a record of collaborators for many years and is constantly added to. The ALSPAC team only removes email addresses from their mailing list if someone explicitly requests this action. Thus, their list likely includes several email addresses that are no longer monitored. For example, we received one email reply stating that the recipient hasn’t been active in research for 30 years. It is also possible that some researchers have multiple email addresses on the mailing list (e.g., because they moved institutions). These two factors may have deflated the response rate.} The median time taken for complete survey responses was 7.40 minutes (IQR: 4.60 to 13.10). This manuscript presents the results for complete surveys. Supplementary Material D presents the results with partially complete surveys included.

@@

Respondents had published a median of NA (IQR 2 to 26.20) studies using preexisting observational data (Supplementary Figure C1). They reported using the following programming languages or software packages: R (n = 65), Stata (n = 48), SPSS (n = 17), SAS (n = 15), Python (n = 6), Mplus (n = 3), Bash (n = 2), MATLAB (n = 1), Nextflow (n = 1), and plink2 (n = 1) (Supplementary Table C1)\footnote[2]{Participants could select multiple responses to this survey question.}. 62\% (62/100) of participants reported being more concerned with research trustworthiness, bias, rigour, and reproducibility compared to what they think of as a typical research who uses preexisting observational data; 6\% (6/100) reported being less concerned (Supplementary Figure C2).

@@

\caption{\textbf{Figure 1. Responses to the survey questions on trustworthiness and reproducibility of observational research with preexisting data and ECAWs.} The survey defined trustworthy as ``meaning that the results and conclusions of the publications are valid, reliable, rigorous, and accurate. That they merit trust''. The survey defined reproducible ``in the sense that other researchers re-analysing the data with the same research question would produce similar results.'' For each item, the number to the left of the data bar indicates the combined percentage for the responses depicted in any shade of brown/orange. The number in the center of the data bar (gray) indicates the percentage of neutral responses. The number to the right of the data bar indicates the combined percentage for the responses depicted in any shade of green. The bar charts in the top panel had no missing responses or selection of the option \emph{``I don't understand the question''}. The bar charts in the bottom panel excluded missing responses (n = ; ) and responses of \emph{``I don't understand the question''} (n = 3; 2).}\label{fig:typicallyEcawPlot}
\end{figure}

@@

\caption{\textbf{Figure 3. Responses to survey questions about using ECAWs.} These bar charts exclude missing values (\emph{n} = 0; 0; 0; 0; respectively from top to bottom), responses of \emph{``I don't understand the question''} (\emph{n} = 0; 4; 1; 1), and responses of \emph{``Unsure''} (\emph{n} = 2; 5; 6; 8). Agreement with the first question may be slightly inflated due to the format of the questions in this block. Respondents with a highly positive inclination towards ECAWs would be expected to disagree with the first question, but agree with the next three questions. Four respondents agreed with all four statements, suggesting they may have glazed over the word ``less'' in the first question \protect\footnotemark[3]. Interpreting responses to the second and third question come with a degree of ambiguity as the survey did not specify what was meant by the term ``study'' \protect\footnotemark[4].}\label{fig:alspacPlot}
\end{figure}

@@

Thankyou very much to those of you who have already completed the survey below.
There is still time to complete the survey (see details in the original email below), which closes of November 1st.

@@

Please send any comments or queries directly to Robert (\href{mailto:robert.thibault@bristol.ac.uk}{\nolinkurl{robert.thibault@bristol.ac.uk}}).
Many thanks,
The ALSPAC Executive

@@

>>>>

Of 1148 emails sent, 1094 went through and 54 bounced. The survey was completed 103 times and partially completed 20 times, leading to a response rate of 9\% for complete surveys and 11\% for at least partially complete surveys.\footnote[1]{The ALSPAC mailing list has been maintained as a record of collaborators for many years and is constantly added to. The ALSPAC team only removes email addresses from their mailing list if someone explicitly requests this action. Thus, their list likely includes several email addresses that are no longer monitored. For example, we received one email reply stating that the recipient hasn’t been active in research for 30 years. It is also possible that some researchers have multiple email addresses on the mailing list (e.g., because they moved institutions). These two factors may have deflated the response rate.} The median time to complete the survey was 7.40 minutes (IQR: 4.60 to 13.10). This manuscript presents the results for complete surveys. Supplementary Material D presents the results with partially complete surveys included.

@@

Respondents had published a median of 10 (IQR 2 to 26) studies using preexisting observational data (Supplementary Figure C1). They reported using the following programming languages or software packages: R (n = 65), Stata (n = 48), SPSS (n = 17), SAS (n = 15), Python (n = 6), Mplus (n = 3), Bash (n = 2), MATLAB (n = 1), Nextflow (n = 1), and plink2 (n = 1) (Supplementary Table C1)\footnote[2]{Participants could select multiple responses to this survey question.}. 62\% (62/100) of participants reported being more concerned with research trustworthiness, bias, rigour, and reproducibility compared to what they think of as a typical research who uses preexisting observational data; 6\% (6/100) reported being less concerned (Supplementary Figure C2).

@@

\caption{\textbf{Figure 1. Responses to the survey questions on trustworthiness and reproducibility of observational research with preexisting data and ECAWs.} The survey defined trustworthy as ``meaning that the results and conclusions of the publications are valid, reliable, rigorous, and accurate. That they merit trust''. The survey defined reproducible ``in the sense that other researchers re-analysing the data with the same research question would produce similar results.'' For each item, the number to the left of the data bar indicates the combined percentage for the responses depicted in any shade of brown/orange. The number in the center of the data bar (gray) indicates the percentage of neutral responses. The number to the right of the data bar indicates the combined percentage for the responses depicted in any shade of green. The bar charts in the top panel had no missing responses or selection of the option \emph{``I don't understand the question''}. The bar charts in the bottom panel excluded responses of \emph{``I don't understand the question''} (n = 3; 2).}\label{fig:typicallyEcawPlot}
\end{figure}

@@

\caption{\textbf{Figure 3. Responses to survey questions about using ECAWs.} These bar charts exclude responses of \emph{``I don't understand the question''} (\emph{n} = 0; 4; 1; 1; respectively from top to bottom), and responses of \emph{``Unsure''} (\emph{n} = 2; 5; 6; 8). Agreement with the first question may be slightly inflated due to the format of the questions in this block. Respondents with a highly positive inclination towards ECAWs would be expected to disagree with the first question, but agree with the next three questions. Four respondents agreed with all four statements, suggesting they may have glazed over the word ``less'' in the first question \protect\footnotemark[3]. Interpreting responses to the second and third question come with a degree of ambiguity as the survey did not specify what was meant by the term ``study'' \protect\footnotemark[4].}\label{fig:alspacPlot}
\end{figure}

@@

Thank you very much to those of you who have already completed the survey below.
There is still time to complete the survey (see details in the original email below), which closes on November 1st.

@@

Many thanks,
The ALSPAC Executive

@@

\hypertarget{participants-2}{%
\subsubsection{Participants}\label{participants-2}}

Respondents had published a median of 10 (IQR 2 to 25) studies using preexisting observational data (Supplementary Figure C1). They reported using the following programming languages or software packages: R (n = 65), Stata (n = 48), SPSS (n = 17), SAS (n = 15), Python (n = 6), Mplus (n = 3), Bash (n = 2), MATLAB (n = 1), Nextflow (n = 1), and plink2 (n = 1) (Supplementary Table C1)\footnote[1]{Participants could select multiple responses to this survey question.}. 61\% (62/101) of participants reported being more concerned with research trustworthiness, bias, rigour, and reproducibility compared to what they think of as a typical research who uses preexisting observational data; 6\% (6/101) reported being less concerned (Supplementary Figure C2).

\hypertarget{survey-results-1}{%
\subsection{Survey results}\label{survey-results-1}}

Most respondents agreed that studies that analyze preexisting observational datasets are trustworthy (71\%; 87/123) and reproducible (77\%; 93/121) (Figure D2, top panel). At the same time, many agreed that a study using an ECAW would be \emph{more} trustworthy (70\%; 70/107) and \emph{more} reproducible (69\%; 74/108) compared to a typical study using preexisting observational data (Figure 1, bottom panel).

{\centering \includegraphics[width=1\linewidth]{figs/typicallyEcawPlotAll-1}

\caption{\textbf{Figure 1D. Responses to the survey questions on trustworthiness and reproducibility of observational research with preexisting data and ECAWs.} The survey defined trustworthy as ``meaning that the results and conclusions of the publications are valid, reliable, rigorous, and accurate. That they merit trust''. The survey defined reproducible ``in the sense that other researchers re-analysing the data with the same research question would produce similar results.'' For each item, the number to the left of the data bar indicates the combined percentage for the responses depicted in any shade of brown/orange. The number in the center of the data bar (gray) indicates the percentage of neutral responses. The number to the right of the data bar indicates the combined percentage for the responses depicted in any shade of green. The bar charts in the top panel had no missing responses or selection of the option \emph{``I don't understand the question''}. The bar charts in the bottom panel excluded missing responses (n = 12; 12) and responses of \emph{``I don't understand the question''} (n = 4; 3).}\label{fig:typicallyEcawPlotAll}
\end{figure}

Over half of respondents reported that their studies using preexisting observational data are preregistered never or almost never (33\%; 41/123), or sometimes (21\%; 26/103) (Figure 2D). About half reported sharing their analysis scripts never or almost never (19\%; 23/123), or sometimes (29\%; 36/123). 69\% (85/123) reported that they never or almost never blind the data analyst. Almost all respondents answered that they use both exploratory (87\%; 107/123) and confirmatory (82\%; 101/123) analyses at least sometimes.

{\centering \includegraphics[width=1\linewidth]{figs/methodPlotAll-1}

\caption{\textbf{Figure 2D. Responses to survey questions about the research practices of participants.}}\label{fig:methodPlotAll}
\end{figure}

25\% (26/106) of respondents agreed (versus 44\%; 47/106 who disagreed) that they would be less willing to use ALSPAC data if they were required to use an ECAW (Figure 3D). 51\% (50/98) agreed (21\%; 21/98 disagreed) that they would opt-in if ALSPAC ran a study on ECAWs. 54\% (54/100) agreed (11\%; 11/100 disagreed) that ALSPAC should run a study on ECAWs. 45\% (45/100) agreed (22\%; 22/100 disagreed) that they would prefer using an ECAW than using typical preregistration.

{\centering \includegraphics[width=1\linewidth]{figs/alspacPlotAll-1}

\caption{\textbf{Figure 3D. Responses to survey questions about using ECAWs.} These bar charts exclude missing values (\emph{n} = 14; 14; 14; 14; respectively from top to bottom), responses of \emph{``I don't understand the question''} (\emph{n} = 0; 5; 2; 1), and responses of \emph{``Unsure''} (\emph{n} = 3; 6; 7; 8). Agreement with the first question may be slightly inflated due to the format of the questions in this block. Respondents with a highly positive inclination towards ECAWs would be expected to disagree with the first question, but agree with the next three questions. Four respondents agreed with all four statements, suggesting they may have glazed over the word ``less'' in the first question \protect\footnotemark[2]. Interpreting responses to the second and third question come with a degree of ambiguity as the survey did not specify what was meant by the term ``study'' \protect\footnotemark[3].}\label{fig:alspacPlotAll}
\end{figure}

\protect\footnotetext[2]{Another four respondents agreed with at least the first and second question, which appear contradictory. We did not preregister these considerations. More careful wording of these questions could have circumvented the ambiguity in interpreting these seemingly contradictory responses.}
\protect\footnotetext[3]{We intended for these questions to ask about ALSPAC running a trial on ECAWs. However, due to the ambiguity around the terms “study”, some respondents may have interpreted this as a survey, focus group, feasibility, or pilot type of study.}
